Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-01 draft.tex
revised version from Bernhard

\section{\uppercase{Introduction}}
\label{sec:Introduction}
Epidemiology is the study of causation of diseases. For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) [??] gather as much information as possible about participants to be assessed towards different diseases.
%%
These information are used to determine risk factors for diseases, help people to improve their lifestyle and their health or to support diagnosis of widespread diseases.
%%
Epidemiological research is strongly hypothesis-driven.
%%
More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
%%
As a result, the hypothesis will be accepted or rejected.
%%

%For epidemiological research, cohort studies are acquired.
For epidemiological research, cohort studies are carried out.
%%
These studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes like gender, age and weight.
%%
Back pain is one of the most frequent diseases in the western civilization. 
%%
Epidemiological analysis of back pain aim to characterize the disease by determining risk factors.
%Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%Since back pain is one of the most frequent diseases in the western civilization, many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, no automatic classification approach to characterize back pain based on lumbar back spine attributes has been established yet.
%%
Furthermore, previous analyses mainly focused on a single parameter or a small subset of parameters [Quelle].
%%

In this paper, we present a thorough analysis of the image-derived data from lumbar back spine datasets and analyze their discriminating power for different characteristics of the disease.
%%
Therefore, we combine data mining algorithms with data visualization techniques.
%%
%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
%%
\com{Da in der Auflistung oefter vom IVA-Framework die Rede ist: Wollen wir hierzu noch was schreiben, z.B. dass das im VIS-Paper definiert wurde und was wir darunter verstehen; sowie dass diese Arbeit als Erweiterung dieses Frameworks gedacht ist?}

@@

\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%\noindent \com{Background. Epidemiological Workflow, focus on statistical resilience, image data hard to analyze due to the large amount of subjects, poor image quality and lack of methods}.
\noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To assess a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as an \emph{confounder} for prostate cancer.
%%
These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
\subsection{Epidemiological Data}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies, analyzing one specific disease and its influences.
%%
We focus on data from large scale cohort studies.
%%
These studies aim to collect as much data as possible for each subject.
%%
As a result, these data can be analyzed regarding many diseases and conditions.
%%

Epidemiological data is heterogenous and incomplete.
%%
For example women-specific questions or data about a disease treatment only for subjects suffering from this condition.
%%
Statistical analysis has to take missing data into account.
%%

@@

Modern cohort studies often comprise medical image data.
%%
These data are hard to analyze, since segmentation algorithms are not generally available and need to be custom-made for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
The lower (\emph{lumbar}) spine is the most stressed spine part.
%%
Lower back pain is one of the most frequent diseases in the western civilization.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, sporting activities and body posture combined with physical stress at work.
%%
The exact causes as well as particularly vulnerable risk groups are not known.
%%
Potential \emph{confounding} effects are also subject of current research.
%%

@@

It aims to trigger \emph{hypothesis generation} by providing visualizations able to concurrently analyzing multiple heterogenous variables at once using correlation measures.

@@

Similar to our work, they quantify image data project it into the non-image data information space.

@@

The plot matrix is well suited for a overview visualization, but takes up much screen space and is therefore only suitable for few variables at once (see Fig.~\ref{fig:image-parameter-range}).

@@

Dai et al. \cite{Dai} defined a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.

@@

Gla{\ss}er et al. \cite{Glasser2013} use variables derived from DCE-MRI image capturing the blood flow in the tumorous tissue.

@@

We combine ideas of both papers by validating the significance of image derived parameter aiming to derive new relationships by training decision trees to explain target variables.

@@

Pinhero et al. \cite{Pinheiro} applied association rule mining to create high-confidence decision rules about the mortality rate in liver cancer data.
%%
Decision rules are easily readable and are well suited for classifying medical data. \com{Das frueher erwaehnen}
%%
The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
%%
Once these rules, get too complex, analyzing these decision rules becomes complex.
%%
Sekhavat and Hoeber \cite{Sekhavat} propose a interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between a initial value and an target value.
%%
The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
%%
\paragraph{SHIP Data Analysis.}
Klemm et al. analyzed the lumbar spine variability of of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier cluster.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
Out data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
%%
Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
%%
Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
%%
\\\\
Unique for our work is the combination of an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in a overview visualization.
%%
\section{The Lumbar Spine Data Set}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows to analyze many variables simultaneously.
%%
Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises of 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}).
%%
%It is an active research area to determine these correlations.
%%

%%
\subsection{Non-Image Data}
%\com{Analysis Using Heterogeneous Correlation}
The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sport activity or nutrition. \com{Ein paar Details hierzu noch - Datentyp, missing data, aufeinander aufbauende parameter}

\subsection{Image Data}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/centerline}
  \caption{
	(a) left: Tetrahedron-based finite element model (FEM) of the lumbar spine, capturing the L1-L5 vertebrae as well as the lumbar spine canal.
	%%
	(a) right: FEM aligned to a MRI data set.
	%%
	The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
	%%
	(b) Using the Frenet formulas, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).
	%%
	The curvature angle ($\alpha$) is calculated using the top, bottom and middle point.
	%%
	We also extracted the variables for each projection axis which to assess their information gain.
	%(b) Using the Frenet formulas, we extract the curvature, torsion and the curvature angle from the lumbar spine canal center line.
	}
  \label{fig:centerline}
\end{figure}
Four trained technicians acquired the medical image data for each subject in a standardized way using a 1.5~Tesla MRI scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
%%
The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
%%

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
%%
The tetrahedron-based finite element model is initialized using three user-placed land marks.
%%
A click on the L3 vertebra center initialized the model, two clicks on the top and bottom of the vertebra describes an initial height estimation.
%%
The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae \cite{Klemm2013VMV} (Fig.~\ref{fig:centerline}~a).
%%
The detection fails for several subjects due imprecise initialization, imaging artifacts or strongly deformed spines.
%%
$2,540$ lumbar spine models were obtained and constitute the foundation of this work.
%%
%%
% \com{From VIS'14 Paper}
% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
% %%
% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
% %%
% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
% %%
% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
% %%
% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
% %%
% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
% %%
% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
% %%
% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
% %%

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%%
In a previous work \cite{Klemm2013VMV,Klemm2014VIS}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
%%
Using the Frenet formulas \cite{Frenet}, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b):
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
%%
	\item \emph{Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

Since men are on average taller than women and people of old age shrink due to bone erosion, gender and age are also confounder (Fig.~\ref{fig:age-gender}).

@@

Since the gender confounder mainly encodes body-height, we decided to divide subjects into body-size groups.

@@

A generalized pairs plot of all image-derived variables \cite{GPLOMS} colored by presence or absence of back pain.

@@

We calculated an association matrix to to assess correlations between the image parameters.

@@

No statistical significant correlation could be observed through all subject groups.

@@

\subsubsection{Assessing the Information Gain Using the PCA}
%%
To determine the information gain per image-derived variables, we calculated a PCA and compared the loadings per dimension.
%%
The first three principal components explain 75\% of variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%% ------------------------------------------
\subsubsection{Heterogenous Correlations}
%%
We then expanded our focus on correlations of image-derived parameters with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables is the data set.
%%
The method uses the following correlation metrics for the different type combination:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
\com{TODO: Abbildung einer Korrelationsmatrix? Waere sehr gross!}
%%
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - identical}.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1 \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter.
%%
Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
%%

@@

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal curvature and torsion.
%%
We rely on predictive classification trained to obtain a complex rule set on how combination of the image-parameter explain non-image variables.
%%
Decision trees are a popular classification method in data mining for creating predictive models.
%%
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
This allows epidemiological domain experts to interpret the results without having deep knowledge about the algorithm creating the tree.
%%
Readability is only granted for small trees, complex structures with many branches are not desirable.
%%
Too many branches also impose overfitting to the data.
\com{Quelle Decision Tree}

\subsection{The C4.5/C5.0 Algorithm}
Using a trained data set, the C4.5 algorithm builds decision trees based on information entropy.
%%
Such a calculation requires a numeric or categorical target variable.
%%
The algorithm then tries to find a decision tree, which divides the samples using the input variables just like the target variables.
%%
This means, that every node in the tree represents the attribute which splits the data most efficient into the target subset.
%%
The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
\begin{algorithm}[]
 Check for base cases\;
 \For{each attribute $a$}{
  Find the normalized information gain ratio from splitting on $a$\;
 }
 Let $a_{best}$ be the attribute with the highest normalized information gain\;
 Create a decision node that splits on $a_{best}$\;
 Recurse on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
 \caption{Building a decision tree using the C4.5 Algorithm}
 \label{alg:c4.5}
\end{algorithm}
C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
%%
We use the \texttt{R} implementation of C5.0 \cite{c5.0lassification}.
%%
% \begin{enumerate}
% %%
% \item Check for base cases
% %%
% \item For each attribute $a$
% %%
% \begin{enumerate}
% %%
% \item Find the normalized information gain ratio from splitting on $a$
% %%
% \end{enumerate}
% %%
% \item Let $a_{best}$ be the attribute with the highest normalized information gain
% %%
% \item Create a decision node that splits on $a_{best}$
% %%
% \item Recurse on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node
% %%
% \end{enumerate}
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
Creating dummy variables bypasses this problem.
%%

@@

This is realized by clicking on a entry on the visualization, which then displays the corresponding decision tree in detail.

@@

They can try out the technique without installing any software.

>>>>

\section{\uppercase{Introduction}}
\label{sec:Introduction}
%%Epidemiology is the study of causation of diseases.
Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) [??] gather as much information as possible about participants to be assessed towards different diseases.
%%
These information are used to determine risk factors for diseases, help people to improve their lifestyle and their health or to support diagnosis of widespread diseases.
%%
Epidemiological research is strongly hypothesis-driven.
%%
More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
%%
As a result, the hypothesis will be accepted or rejected.
%%

%For epidemiological research, cohort studies are acquired.
For epidemiological research, cohort studies are carried out.
%%
Since recently, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as gender, age and weight.
%%
Back pain is one of the most frequent diseases in the western civilization. 
%%
Epidemiological analysis of back pain aims to characterize the disease by determining risk factors.
%Many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%Since back pain is one of the most frequent diseases in the western civilization, many epidemiological-motivated approaches aim at improved characterization of this disease with respect to patient-specific attributes.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for the presence of back pain, no automatic classification approach to characterize back pain based on lumbar back spine attributes has been established.
%%
Furthermore, previous analyses mainly focused on a single parameter or a small subset of parameters [Quelle].
%%

In this paper, we present a thorough analysis of the image-derived data from lumbar back spine datasets and analyze their discriminating power for different characteristics of the disease.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
%%
\com{Da in der Auflistung oefter vom IVA-Framework die Rede ist: Wollen wir hierzu noch was schreiben, z.B. dass das im VIS-Paper definiert wurde und was wir darunter verstehen; sowie dass diese Arbeit als Erweiterung dieses Frameworks gedacht ist?}

@@

\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%\noindent \com{Background. Epidemiological Workflow, focus on statistical resilience, image data hard to analyze due to the large amount of subjects, poor image quality and lack of methods}.
\noindent Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To assess a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{confounder} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
\subsection{Epidemiological Data}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies, analyzing one specific disease and its influences on the human body.
%%
We focus on data from large scale cohort studies.
%%
These studies aim to collect as much data as possible for each subject.
%%
As a result, these data can be analyzed regarding many diseases and conditions.
%%

Epidemiological data is heterogenous and incomplete.
%%
For example women-specific questions or data about a disease treatment only subjects suffering from this condition.
%%
Statistical analysis has to take missing data into account.
%%

@@

Modern cohort studies often comprise medical image data.
%%
These data are hard to analyze.
%%
Individual diagnosis or manual segmentation of each body structure by an radiologists is tedious, costly and comprises little reproducibility.
%%
Segmentation algorithms are not generally available and need to be custom-made for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
The lower (\emph{lumbar}) spine is the most stressed spine part.
%%
Lower back pain is one of the most frequent diseases in the western civilization.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, sporting activities and body posture combined with physical stress at work.
%%
The exact causes as well as particularly vulnerable risk groups are not known.
%%
Potential \emph{confounding} effects are also subject of current research.
%%

@@

It aims to trigger \emph{hypothesis generation} by providing visualizations able to concurrently analyze multiple heterogenous variables at once using correlation measures.

@@

Similar to our work, they quantify image data and project it into the non-image data information space.

@@

The plot matrix is well suited for an overview visualization, but takes up much screen space and is therefore only suitable for few variables at once (see Fig.~\ref{fig:image-parameter-range}).

@@

Dai and Gahegan \cite{Dai} defined a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.

@@

Gla{\ss}er et al. \cite{Glasser2013} use variables derived from DCE-MRI data capturing the perfusion in the tumorous tissue.

@@

We combine ideas of both papers by validating the significance of image-derived parameter aiming to derive new relationships by training decision trees to explain target variables.

@@

Pinhero et al. \cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables such as age, gender and geographic location about the mortality rate in liver cancer data.
%%
Decision rules are easily readable and are well suited for classifying medical data. \com{Das frueher erwaehnen}
%%
The results can be easily communicated to domain experts with little knowledge about the algorithm behind the results.
%%
Once these rules, get too complex, analyzing these decision rules becomes complex.
%%
Sekhavat and Hoeber \cite{Sekhavat} propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial value and an target value.
%%
The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
%%
\paragraph{SHIP Data Analysis.}
Klemm et al. analyzed the lumbar spine variability of of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier cluster.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
The data extraction process is based on the detection algorithm and lumbar spine canal extraction presented in this paper.
%%
Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
%%
%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
%%
\\\\
Unique for our work is the combination of an IVA-approach by observing interesting feature relations in context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in an overview visualization.
%%
\section{The Lumbar Spine Data Set}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows to analyze many variables simultaneously.
%%
Therefore, epidemiologists compiled the data set with a with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises of 6.753 subjects from two cohorts (4.420 from \texttt{SHIP-Trend-0} and 2.333 from \texttt{SHIP-2}).
%%
%It is an active research area to determine these correlations.
%%

%%
\subsection{Non-Image Data}
%\com{Analysis Using Heterogeneous Correlation}
The variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors as sporting activity or nutrition. \com{Ein paar Details hierzu noch - Datentyp, missing data, aufeinander aufbauende parameter}

\subsection{Image Data}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/centerline}
  \caption{
	(a) left: Tetrahedron-based finite element model (FEM) of the lumbar spine, capturing the L1-L5 vertebrae as well as the lumbar spine canal.
	%%
	(a) right: FEM aligned to a MRI data set.
	%%
	The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
	%%
	(b) Using the Frenet formulas, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).
	%%
	The curvature angle ($\alpha$) is calculated using the top, bottom and middle point.
	%%
	We also extracted the variables for each projection axis to assess their information gain.
	%(b) Using the Frenet formulas, we extract the curvature, torsion and the curvature angle from the lumbar spine canal center line.
	}
  \label{fig:centerline}
\end{figure}
Four trained radiology technicians acquired the medical image data for each subject in a standardized way using a 1.5~Tesla MRI scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
%%
The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
%%

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
%%
The tetrahedron-based finite element model is initialized using three user-placed landmarks.
%%
A click on the L3 vertebra center initialized the model, two clicks on the top and bottom of the vertebra describes an initial height estimation.
%%
The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae \cite{Klemm2013VMV} (Fig.~\ref{fig:centerline}~a).
%%
The detection fails for several subjects due imprecise initialization, imaging artifacts or strongly deformed spines.
%%
$2,540$ lumbar spine models were obtained and constitute the foundation of this work.
%%
%%
% \com{From VIS'14 Paper}
% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
% %%
% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
% %%
% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
% %%
% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
% %%
% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
% %%
% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
% %%
% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
% %%
% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
% %%

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%%
In a previous work \cite{Klemm2013VMV}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
%%
Using the Frenet formulas \cite{Frenet}, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b):
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
%%
	\item \emph{Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

Since men are on average taller than women and people of old age shrink due to bone erosion, gender and age are also confounders (Fig.~\ref{fig:age-gender}).

@@

Since the gender confounder mainly encodes body-size, we decided to divide subjects into body-size groups.

@@

A generalized pairs plot of all image-derived variables colored by presence or absence of back pain.

@@

We calculated an association matrix to assess correlations between the image parameters.

@@

No statistically significant correlation could be observed through all subject groups.

@@

\subsubsection{Assessing the Information Gain Using the PCA}
%%
To determine the information gain per image-derived variables, we calculated a PCA and compared the loadings per dimension.
%%
The first three principal components explain 75\% of the variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of the variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of the variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%% ------------------------------------------
\subsubsection{Heterogenous Correlations}
%%
We then expanded our focus on correlations of image-derived parameters with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables is the data set.
%%
The method uses the following correlation metrics for the different type combination:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
\com{TODO: Abbildung einer Korrelationsmatrix? Waere sehr gross!}
%%
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - identical}.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1 \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter.
%%
Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
%%

@@

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal curvature and torsion.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-parameter explain non-image variables.
%%
Decision trees are a popular classification method in data mining for creating predictive models.
%%
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
This allows epidemiologists to interpret the results without having deep knowledge about the algorithm creating the tree.
%%
Readability is only granted for small trees, complex structures with many branches are not desirable.
%%
Too many branches also impose overfitting to the data.
\com{Quelle Decision Tree}

\subsection{The C4.5/C5.0 Algorithm}
Using a trained data set, the C4.5 algorithm builds decision trees based on information entropy.
%%
Such a calculation requires a numeric or categorical target variable.
%%
The algorithm then tries to find a decision tree, which divides the samples using the input variables just like the target variables.
%%
This means, that every node in the tree represents the attribute which splits the data most efficiently into the target subset.
%%
The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
\begin{algorithm}[]
 Check for base cases\;
 \For{each attribute $a$}{
  Find the normalized information gain ratio from splitting on $a$\;
 }
 Let $a_{best}$ be the attribute with the highest normalized information gain\;
 Create a decision node that splits on $a_{best}$\;
 Recurse on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
 \caption{Building a decision tree using the C4.5 Algorithm}
 \label{alg:c4.5}
\end{algorithm}
C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
%%
We use the \texttt{R} implementation of C5.0 \cite{c5.0lassification}.
%%
% \begin{enumerate}
% %%
% \item Check for base cases
% %%
% \item For each attribute $a$
% %%
% \begin{enumerate}
% %%
% \item Find the normalized information gain ratio from splitting on $a$
% %%
% \end{enumerate}
% %%
% \item Let $a_{best}$ be the attribute with the highest normalized information gain
% %%
% \item Create a decision node that splits on $a_{best}$
% %%
% \item Recurse on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node
% %%
% \end{enumerate}
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
Creating dummy variables bypasses this problem.
%%

@@

This is realized by clicking on an entry on the visualization, which then displays the corresponding decision tree in detail.

@@

They can use the technique without installing any software.
